Bayesian Computational Strategies and Implementations

### 2021-04-28 Paper.Rmd
add updated paper and simulations

---
title: "Bayes Computational Methods"
author: "Evan Miyakawa"
output: pdf_document
---

@@

@@

In Bayesian statistical inference, we are interested in finding the posterior distribution of certain parameters, given a prior distribution and a likelihood for those parameters our data. Much of Bayesian theory is built upon the definition of conditional probability, which states that we can find the probability of a variable $\theta$, conditioned on $x$, by $$ p(\theta | x) = \frac{p(x | \theta) p(\theta)}{p(x)}.$$ In any Bayesian inference, the first step is to find the joint probability distribution of all the observed and unobserved quantities in the problem. Using the definition of conditional probability, we can then find the posterior distribution of the parameters of interest. Bayes theorem states that the posterior distribution of a parameter $\theta$ is equal to $$p(\theta | x_1, x_2, \dots, x_n) = \frac{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta)}{\int{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta) d \theta}} = \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}},$$ where $\ell(\theta | \textbf{x})$ is proportional to the likelihood of $\theta$, conditioned on the data, and $\pi(\theta)$ is the prior distribution of $\theta$.

For example, suppose that $X_i \sim \text{Bern}(\theta)$, and we are interested in inference on $\theta$. A reasonable prior to assign to $\theta$ would be a $\text{Beta}(\alpha, \beta)$, distribution, given that $\theta$ is constrained to be between 0 and 1, which is a property of the Beta distribution. Therefore, given that $\theta \sim \text{Beta}(\alpha, \beta)$ for some known values of $\alpha$ and $\beta$, then 
\begin{align*}
  p(\theta | x) &= \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}} \\
  &= \frac{\theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1}}{\int \theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1} d \theta} \\
  &\propto \theta ^ {\sum (x_i) + \alpha - 1} (1 - \theta) ^ {n - \sum (x_i) + \beta - 1},
\end{align*}
which is the kernel of a $\text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta \right)$ distribution. Therefore, we can analytically calculate the posterior distribution of $\theta$ as $$p(\theta | x) \sim \text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta\right).$$

@@

The need for bayesian computational methods arises in our search for an alternative approach to finding the posterior if we cannot do so analytically. Most of the time, the goal of any bayesian analysis is to not to find the posterior distribution, but to calculate expectations with respect to the posterior. However, in order for these estimates to be accurate, we often need to know about the full behavior of that posterior distribution. In bayesian computing, the aim is to somehow obtain individual draws of values that would come from the posterior, and then aggregate a lot of them to compute the expectations we need.

@@

Markov Chain Monte Carlo (referred to as MCMC), is a common approach that allows for drawing samples from a target distribution, which in our case is the posterior. MCMC is an iterative algorithm, where we obtain one draw at a time from the parameter space from which we are trying to sample. In any Markov chain, the marginal distribution of each draw is conditioned on the values of our chain from the previous draw. As we keep obtaining draws, these marginal distributions asymptotically converge to the true target distribution. Therefore, if we run the markove chain for many iterations, we can expect to see draws coming from the true posterior distribution we are trying to estimate. MCMC is the only suitable method for obtaining the posterior distribution when we don't have conjugate priors.

@@

When using Markov Chain methods to sample from the Bayesian posterior distribution, the Metropolis-Hastings Algorithm gives a general approach to construct transitions from one draw to the next. The algorithm has two steps: a proposal step, where we suggest a new draw, and a correction step, where we try to reject any proposals that move too far away from the typical set from which we are trying to sample (Betancourt paper).

@@

There are some issues that arise with the Metropolis-Hastings algorithm. Each proposal is biased towards larger volumes (such as the tails of the distribution). Although the algorithm will reject these proposals because the density is smaller, this can lead to many rejected proposals. This is especially true in higher dimensions, where in some instances, virtually every proposal step will be rejected, and even finishing one "journey" across the typical set might be unattainable. In better scenarios, there is still often autocorrelation in the draws that has to be accounted for.

@@

Gibbs Sampling is one of the most common Markov Chain Monte Carlo Algorithms used for Bayesian computation. A special case of the Metropolis-Hastings algorithm, a Gibbs sampler works by generating a multi-dimensional markov chain, with one dimension for each unknown variable of interest. By splitting the vector of random unknown variables, the algorithm samples each subvector one by one, conditional on the most recent values of the other variables.

@@

    $$\theta_1^{(1)} \sim p(\theta_1 | \theta_2^{(0)}, \theta_3^{(0)}, \dots , \theta_k^{(0)}, y).$$
    - Sample a new value $\theta_2^{(1)}$ for the second component of $\theta$, from the full conditional distribution $p(\theta_2 | \theta_1^{(1)}, \theta_3^{(0)}, \dots , \theta_k^{(0)}, y)$.
    - Continue this process for the remainder of the $k$ elements of $\theta$.
3. Repeat step 2 for the total decided number of iterations that the sampler will run.

@@

A main of objective of Hamiltonian Monte Carlo methods is to use our knowledge of the geometry of the target distribution to guide our sampling efforts, instead of utilizing a random-walk algorithm (Betancourt HMC Youtube talk). The density function of the target distribution is translated into a potential energy function, which is used in Hamiltonian Dynamics, a physics system that we employ in order to yield monte carlo draws. To use this dynamic system to obtain draws, each parameter is transformed into a position variable, where the position of the variable can be interpreted as the value of the parameter at that point in time. Each position variable $q$ also has one associated momentum variable $p$ which will guide the movement of the position variable inside the simulation.

@@

We can simulate a Hamiltonian monte carlo markov chain where each iteration consists of the following steps:

@@

![\textit{A visual representation from Betancourt of lifting up off the parameter space (illustrated in red) to a higher-dimension phase space to sample when using HMC.}](Paper-Figures/Blank.png)

While the probability distribution for the position variables is defined by the target distribution, the choice of the conditional distribution the momentum, which is equivalent the kinetic energy function, is up to the user. This choice is very important as it can drastically affect the simulation. There are several common gaussian distributions used for these distributions, which are used by default in the implementations that utilize HMC methods.

Hamiltonian Monte Carlo methods are very robust, in that the markov chains get "stuck" in complex regions of the posterior less frequently than other common MCMC methods. When the markov chains do get stuck, it is easy for the program running the sampler to spot and report to the user.

# Bayesian Samplers

### BUGS

The BUGS project, which stands for Bayesian inference Using Gibbs Sampling, began in 1989 in an effort to create a program that utilized MCMC to approximate draws from the posterior distribution in a Bayesian framework. The BUGS language has functions and distributions that allow for specifying relationship between different "nodes" in a bayesian model. While the language itself has a somewhat limited syntax, it is able to describe a very diverse and complex set of models. The BUGS language is declarative, meaning that the order of statements in a model script does not matter since the entire script is evaluated at once, instead of line-by-line in a procedural language.

Many different programs have been developed that utilize the BUGS syntax to accomplish bayesian sampling, and while these programs have continually evolved, the BUGS syntax has remained very consistent. The Windows program WinBUGS was created by the inventors of the BUGS language, and was a more popular choice for bayesian computation until it became deprecated in 2007. OpenBUGS is now more commonly used in place of WinBUGS, although it's last stable release was in 2014. OpenBUGS communicates better with other software than WinBUGS and is more portable.

In general, there are six steps needed to run bayesian models in WinBUGS/OpenBUGS:

1. Specify the model through BUGS syntax and check to make sure the model code is valid.
2. Load the appropriate data.
3. Compiling the BUGS model.
4. Start the simulation at appropriate initial values for the Markov chains.
5. Simulate draws from the posterior distributions of the unknown parameters of interest.
6. Gather the results and report summary statistics, along with other appropriate output.

@@

OpenBUGS is the oldest of the Bayesian samplers that we use, maintained only as recently as 2014. OpenBugs is a program for Windows machines, but there are workarounds to run it on MacOS and Linux, such as WINE or a virtual machine.  OpenBUGS can be used through the actual click-and-point application, or through \texttt{R}, using the \texttt{R2OpenBugs} package, which allows for the user to send model scripts and data to OpenBugs without needing to leave the \texttt{R} interface. OpenBUGS is based on the BUGS model specification language, which is a small-to-moderately sized set of functions and commands that allow for extensive specification of bayesian models. Several other common bayesian sampling implementations are made to closely mimic the BUGS language as well.

<!-- Using this method causes OpenBugs to run invisibly in the background and return the results, unless an argument is specified to debug code in the actual application itself, which can be very useful. When using R to run OpenBugs, there is little way to know what went wrong if the code errors somewhere without interacting with the OpenBugs program itself. Using the \texttt{debug = T} argument within \texttt{bugs()} will cause the application to run in the forefront as if it were actually being used directly by the user. This way, any messages given by the application while running can be viewed directly. It is recommend to use \texttt{debug = T} only when debugging, as it does slow down the sampling process slightly.  -->

OpenBugs has many tools to visualize the performance the sampler which are easy to use. The biggest drawback of using the OpenBugs application directly is that specifiying the data can be tedious if the data is not very small in size or design. Using R to specify the data is much easier. One unique feature in OpenBugs is the Doodle tool, which allows for model specification through the creation of a diagram of nodes representing the parameters. Though using Doodle can be helpful for illustration, it is not used often in daily practice.

@@

@@

In terms of model specification, JAGS is very similar to OpenBugs, as it is based on the BUGS model specification language. However, JAGS is always interacted with entirely in R, making the bayesian inference process more seemless, using the R package \texttt{rjags}. The \texttt{runjags} is an optional package that allows for the JAGS model script to be inputted as a string, and run using the \texttt{run.jags()} function. Built on C++, JAGS is able to run MCMC sampling process very quickly. There is plenty of documentation and support available for JAGS, which is still maintained.

@@

Nimble allows for bayesian computation within \texttt{R}, like JAGS and OpenBUGS, but is also made for more flexible statistical modeling. It uses C++ under the hood to compile models and algorithms. Using the \texttt{nimble} package, Nimble can perform MCMC bayesian sampling using common algorithms, but also allows for other types of algorithms to be loaded or custom-built. It can also compile other code without BUGS models into C++ for other uses. For the purpose of bayesian sampling using MCMC, Nimble is used very similarly to JAGS and OpenBUGS in practicality, as its model specification language is similar. The flexibility of Nimble is not needed in most cases when doing bayesian MCMC methods. The default sampling method in Nimble is Gibbs sampling, but other methods can be used as well. Because the model has to be compiled first for C++ before sampling each time, this can drastically increase the total computation time as compared to other sampling methods.

## STAN

STAN is a language in-and-of itself used for statistcal modeling and data analysis. STAN can be used in several programming languages, including \texttt{R} and Python. STAN uses Hamiltonian Monte Carlo sampling methods, which ideally can reduce the computational time for many types of statistical models. Like Nimble, STAN compiles each model for C++, but one convenient feature is that the user can store each compiled model in a file, which allows for repeated use without the need to re-compile each time. In \texttt{R}, STAN is used through the \texttt{rstan} package, which allows for data to be loaded through normal \texttt{R} data structures. The biggest practical difference in using STAN compared to other bayesian computational methods is in the model specification process. STAN requires a model object in the form of a .stan file, which is created in a very unique language that is more formal than BUGS syntax. A .stan file, either created through a character string in R or in a seperate text file, can be used by the \texttt{stan()} function, which takes the file along with the data to perform the sampling. There is a very robust set of manuals for STAN users, including many papers written on the subject.

## Greta

Greta, a more recent addition to the list of bayesian computational implementations, is made specifically for easy use in \texttt{R}. The model and data specification process is intended to be as straightforward as possible users who are familiar with \texttt{R} but not as familiar with other bayes methods. Not only does it utilize Hamiltonian Monte Carlo methods for sampling, but it also uses Google TensorFlow,a machine learning platform from Google, to perform its calculations. This allows for more flexible use of CPUs and GPUs. Greta offers a model visualization tool that does not exist in any of the other Bayesian samplers, in that it can provide a graphical representation of the specified model through the use of nodes and arrows. This tool can be very useful in trying to explain how all the components of a model work together.

Unfortunately, as of the writing of this, Greta is still in need of lots of development, and lacks an appropriate amount of documentation and support for users, compared to the other implementations mentioned previously. The sampling process can slow compared to the other bayes methods when using identical sampling configurations, especially for simple models. There are also several types of models that Greta seems unable to properly specify at the moment.

@@

@@

model {
  y ~ dbin(p,n)
  p ~ dbeta(1,1)
}

##
```

@@

@@

@@

## STAN

Running models in STAN requires a .stan file which contains the model script code, written in the unique STAN syntax. This can either be a seperate file, or can be inputted as a character string. It is generally recommended to use a seperate .stan file, espcially when using RStudio, which can do syntax highlighting and point out errors in STAN code.

Here is the .stan file for this example, which is saved as "stan_file.stan". Each STAN file contains a data block, parameters block, and model block at a minimum. The data block initializes each data variable, the parameters block initializes any parameters to monitor, and the model block contains more familiar looking model script code.

@@

model {
  y ~ binomial(n,p);
  p ~ beta(1,1);
}

##
```

@@

## Greta

Model specification in Greta is more intuitive for R users, but requires the extra step of converting existing data objects into Greta arrays, using the `as_data()` function from the `greta` package. 
```{r}
## convert data into greta arrays
y <- greta::as_data(y)
n <- greta::as_data(n)
```

@@

When using Markov Chain Monte Carlo methods to estimate Bayesian posteriors of parameters, the number of independent Markov chains to be run needs to be specified. It is recommended to use several chains to explore the space of the target distribution, as a single chain may not fully traverse the entire region on its own. For some simpler models that don't have irregular features in the target distribution, using one chain can suffice. Each sampler has an argument in its function call that specifies the number of chains. Some of the samplers have a default value if the number of chains isn't given.

@@

The number of steps that each Markov chain will take is a very important setting to get right. If each chain is not set up to run long enough, the parameter space may not be fully explored and the estimates will be inaccurate. Checking convergence statistics can be helpful in determining if the number of iterations is sufficient. An argument specifiying the number of iterations for each chain can be inputted in the function call for each sampler.

@@

Since our inferences are based on the assmumption that our draws are from the target distribution, it is wise to discard the samples at the beginning of each chain, since it takes time for the sampler to "settle down". The terms "warmup" and "burn-in" are often used to indicate the number of iterations that we will discard at the beginning of each chain when calculating statistics about our posterior distribution. We can choose this number in the function call for each sampler. In \texttt{OpenBUGS}, \texttt{Nimble}, and \texttt{Jags}, the argument is referred to as burnin, but in \texttt{STAN} and \texttt{greta} it is called warmup. Some samplers (\texttt{greta}, \texttt{Jags, Nimble}) treat this warmup argument as a number of iterations to run before starting to track the sampling results, but others (\texttt{STAN}, \texttt{BUGS}) have this be the actual number of iterations to discard out of the total number of specified iterations, so it is important to know which designation a sampler uses.

@@

When describing a bayesian model through code, there are instances were slightly different implementations are needed based on the sampler being used. Both in model and data specification, the user needs to understand the intricacies of the specific sampling tool being used in order to get the analysis to run properly. This section provides some helpful tips for implementing models in each of the different samplers being examined.

>>>>

---
title: "Bayesian Computational Strategies and Implementations"
author: "Evan Miyakawa"
output: pdf_document
---

@@

\textcolor{red}{Introductory paragraph.}

@@

In Bayesian statistical inference, we are interested in finding the posterior probability distribution of certain parameters, given a prior distribution and a likelihood for those parameters in our data. Much of Bayesian theory is built upon the definition of conditional probability, which states that we can find the probability of a variable $\theta$, conditioned on $x$, by $$ p(\theta | x) = \frac{p(x | \theta) p(\theta)}{p(x)}.$$ In any Bayesian inference, the first step is to find the joint probability distribution of all the observed and unobserved quantities in the problem. Using the definition of conditional probability, we can then find the posterior distribution of the parameters of interest. Bayes theorem states that the posterior distribution of a parameter $\theta$ is equal to $$p(\theta | \textbf{x}) = \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}},$$ where $\ell(\theta | \textbf{x})$ is proportional to the likelihood of $\theta$, conditioned on the data $\textbf{x}$, and $\pi(\theta)$ is the prior distribution of $\theta$.

<!-- Bayes theorem states that the posterior distribution of a parameter $\theta$ is equal to $$p(\theta | x_1, x_2, \dots, x_n) = \frac{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta)}{\int{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta) d \theta}} = \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}},$$ where $\ell(\theta | \textbf{x})$ is proportional to the likelihood of $\theta$, conditioned on the data, and $\pi(\theta)$ is the prior distribution of $\theta$. -->

For example, suppose that $X_i \sim \text{Bern}(\theta)$, and we are interested in inference on $\theta$. A common prior distribution for $\theta$ would be a $\text{Beta}(\alpha, \beta)$, given that $\theta$ is constrained to be between 0 and 1, which is a property of the Beta distribution. Therefore, given that $\theta \sim \text{Beta}(\alpha, \beta)$ for some known values of $\alpha$ and $\beta$, then 
\begin{align*}
  p(\theta | x) &= \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}} \\
  &= \frac{\theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1}}{\int \theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1} d \theta} \\
  &\propto \theta ^ {\sum (x_i) + \alpha - 1} (1 - \theta) ^ {n - \sum (x_i) + \beta - 1},
\end{align*}
which is the kernel of a $\text{Beta} \left( \sum (x_i) + \alpha, n - \sum (x_i) + \beta \right)$ distribution. Therefore, we can analytically calculate the posterior distribution of $\theta$ as $$p(\theta | x) \sim \text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta\right).$$

@@

The need for bayesian computational methods arises in our search for an alternative approach to finding the posterior if we cannot do so analytically. Most of the time, the goal of any bayesian analysis is to not to find the posterior distribution, but to calculate expectations with respect to the posterior. However, in order for these estimates to be accurate, we often need to know about the full behavior of that posterior distribution. In bayesian computing, the aim is to obtain individual draws of values that would come from the posterior, and then aggregate a large number of them to compute the expectations we need.

@@

Approaches using Markov Chain Monte Carlo methods (referred to as MCMC), are commonly used to drawing samples from a target distribution, which in our case is the posterior distribution of our parameter(s). MCMC is an iterative algorithm, where we obtain one draw at a time from the parameter space from which we are trying to sample. In any Markov chain, the marginal distribution of each draw is conditioned on the values of our chain from the previous draw. As we keep obtaining draws, these marginal distributions asymptotically converge to the true target distribution. Therefore, if we run the markov chain for many iterations, we can expect to see draws coming from the true posterior distribution we are trying to estimate. MCMC is the only suitable method for obtaining the posterior distribution when we don't have conjugate priors.

@@

When using Markov Chain methods to sample from a posterior distribution, the Metropolis-Hastings Algorithm gives a general approach to construct transitions from one draw to the next. The algorithm has two steps: a proposal step, where we suggest a new draw, and a correction step, where we potentially reject a proposal that move too far away from the typical set from which we are trying to sample (Betancourt paper).

@@

There are some issues that arise with the Metropolis-Hastings algorithm. Each proposal is biased towards larger volumes (such as the tails of the distribution). Although the algorithm will reject these proposals because the density is smaller, this can lead to many rejected proposals. This is especially true in higher dimensions, where in some instances, virtually every proposal step will be rejected, and finishing just one "journey" across the typical set might be unattainable. In simpler scenarios, there is still often autocorrelation in the draws for which we must account (Betancourt).

@@

Gibbs Sampling is one of the most common Markov Chain Monte Carlo algorithms used for Bayesian computation. A special case of the Metropolis-Hastings algorithm, a Gibbs sampler works by generating a multi-dimensional markov chain, with one dimension for each unknown variable of interest. By splitting the vector of random unknown variables, the algorithm samples each subvector one by one, conditional on the most recent values of the other variables.

@@

    $$\theta_1^{(1)} \sim p(\theta_1 | \theta_2^{(0)}, \theta_3^{(0)}, \dots , \theta_k^{(0)}, y).$$
    - Sample a new value $\theta_2^{(1)}$ for the second component of $\theta$, from the full conditional distribution $p(\theta_2 | \theta_1^{(1)}, \theta_3^{(0)}, \dots , \theta_k^{(0)}, y)$.
    - Continue this process for the remainder of the $k$ elements of $\theta$.
3. Repeat step 2 for the total decided number of iterations that the sampler will run (BDA3).

@@

The main of objective of Hamiltonian Monte Carlo (HMC) algorithms is to use the knowledge of the geometry of the target distribution to guide sampling efforts, instead of using a random-walk algorithm (Betancourt HMC Youtube talk). Under HMC, the density function of the target distribution is translated into a potential energy function, which is used in Hamiltonian Dynamics, a physics system that is employed to yield Monte Carlo draws. To use this dynamic system to obtain draws, each parameter is transformed into a position variable, where the position of the variable can be interpreted as the value of the parameter at that point in time. Each position variable $q$ also has one associated momentum variable $p$ which will guide the movement of the position variable inside the simulation.

@@

We can simulate a Hamiltonian Monte Carlo Markov chain where each iteration consists of the following steps:

@@

![\textit{A visual representation from Betancourt of lifting up off the parameter space (illustrated in red) to a higher-dimension phase space to sample when using HMC (Betancourt).}](Paper-Figures/Blank.png)

While the probability distribution for the position variables is defined by the target distribution, the choice of the conditional distribution of the momentum, which is equivalent the kinetic energy function, is up to the user. This choice is very important as it can drastically affect the simulation. There are several common gaussian distributions used for these distributions, which are used by default in the implementations that employ HMC methods (Betancourt).

Hamiltonian Monte Carlo algorithms are very robust, in that the markov chains get "stuck" in complex regions of the posterior less frequently than other common MCMC methods. When the Markov chains do get stuck, it is easy for the program running the sampler to spot the issues and report them to the user.

There are some inherent challenges that come with HMC methods that are worth mentioning. In every transition step between MCMC draws, a system of differential equations needs to be solved in order to direct the movement of the position variables. This can be difficult in some situations. Additionally, getting the "physics simulation" set up properly for a given model and data set requires several tuning parameters to be set correctly. While some of these parameters can be picked automatically through an adaptation phase before sampling, the user needs to have some knowledge of how to set these initially. This is less of an issue for the no-U-turn sampler, an HMC algorithm that can automatically tune the parameters in a warmup and adaptation phase that occurs before sampling (Gelman, Intro to Stan).

# Platforms for Bayesian Sampling and Inference

The term "probabilistic programming language" refers to programming languages and tools that natively work with probabilistic models and do inference on them more generally (Hardesty). Within bayesian computing, there are several platforms that are used to do MCMC sampling, to great effect. A bayesian computing platform usually consists of a specification language used to write and program the models, and the engine that performs the computations. We have chosen several of the most common of these platforms to look at more in detail in this paper, with the objective of portraying the strengths, weaknesses, and unique features of each.

<!-- ### BUGS -->

<!-- The BUGS project, which stands for Bayesian inference Using Gibbs Sampling, began in 1989 in an effort to create a program that utilized MCMC to approximate draws from the posterior distribution in a Bayesian framework. The BUGS language has functions and distributions that allow for specifying relationship between different "nodes" in a bayesian model. While the language itself has a somewhat limited syntax, it is able to describe a very diverse and complex set of models. The BUGS language is declarative, meaning that the order of statements in a model script does not matter since the entire script is evaluated at once, instead of line-by-line in a procedural language.  -->

<!-- Many different programs have been developed that utilize the BUGS syntax to accomplish bayesian sampling, and while these programs have continually evolved, the BUGS syntax has remained very consistent. The Windows program WinBUGS was created by the inventors of the BUGS language, and was a more popular choice for bayesian computation until it became deprecated in 2007. OpenBUGS is now more commonly used in place of WinBUGS, although it's last stable release was in 2014. OpenBUGS communicates better with other software than WinBUGS and is more portable.  -->

<!-- In general, there are six steps needed to run bayesian models in WinBUGS/OpenBUGS: -->

<!-- 1. Specify the model through BUGS syntax and check to make sure the model code is valid. -->
<!-- 2. Load the appropriate data. -->
<!-- 3. Compiling the BUGS model. -->
<!-- 4. Start the simulation at appropriate initial values for the Markov chains. -->
<!-- 5. Simulate draws from the posterior distributions of the unknown parameters of interest. -->
<!-- 6. Gather the results and report summary statistics, along with other appropriate output. -->

@@

OpenBUGS is the oldest of the Bayesian sampling platforms that we use, maintained only as recently as 2014. OpenBUGS is a program for Windows machines, but there are workarounds to run it on MacOS and Linux, such as WINE or a virtual machine.  OpenBUGS can be used through the actual click-and-point application, or through R, using the `R2OpenBUGS` package, which allows for the user to send model scripts and data to OpenBUGS without needing to leave the R interface. OpenBUGS is based on the BUGS model specification language, which is a small-to-moderately sized set of functions and commands that allow for extensive specification of bayesian models. The BUGS language is declarative, meaning that order in which expressions are listed in the model code does not matter (Lunn, BUGS Project).

The program uses Gibbs sampling methods by default, but also allows for Metropolis-Hastings and Slice sampling. Depending on the model type, OpenBUGS will attempt to group nodes that are correlated together and update them at the same time in each Gibbs sampling transition. In most cases this is not feasible, and parameter nodes are updated one at a time, which can be very slow if there are many parameters or there is a high level of correlation between them (OpenBUGS Manual).

<!-- Using this method causes OpenBUGS to run invisibly in the background and return the results, unless an argument is specified to debug code in the actual application itself, which can be very useful. When using R to run OpenBUGS, there is little way to know what went wrong if the code errors somewhere without interacting with the OpenBUGS program itself. Using the \texttt{debug = T} argument within \texttt{bugs()} will cause the application to run in the forefront as if it were actually being used directly by the user. This way, any messages given by the application while running can be viewed directly. It is recommend to use \texttt{debug = T} only when debugging, as it does slow down the sampling process slightly.  -->

OpenBUGS has tools to visualize the performance diagnostics of the sampling, which are easy to use. The biggest drawback of using the OpenBUGS application directly is that specifying the data by hand can be tedious if the data is not very small in size or design structure. Using R to specify the data is much easier.

One main purpose of the BUGS language and software is to easily work with directed acyclical graphs (DAGs). The syntax makes specifying child-parent relationships easy, and the program can interpret the model code and automatically determine the nodes relevant to each full conditional distribution calculation used in Gibbs sampling. Because of how flexible the BUGS language is, any DAG can be represented through a BUGS model (Lunn). One of the unique feature in OpenBUGS is the Doodle tool, which allows for model specification through the creation of a DAG itself by the user, where each represents a variable. A Doodle can be converted into a model, though this is not used often in daily practice.

@@

There are a few well-known weaknesses to OpenBUGS. The software only works on Windows platforms, although there have been ways to use the program on MacOS through The program itself can be prone to failure on occasion, and there are no built-in diagnostic checks to alert the user if the sampling has gone poorly. Diagnosing errors in the model code can be somewhat tedious if the errors given aren't very descriptive. OpenBUGS does not support logical expressions such as the if-else statement, although there are workarounds, such as the `step()` function that is built-in (Lunn).

@@

JAGS (Just Another Gibbs Sampler) is similar in purpose to OpenBUGS, in that it does analysis of Bayesian models using MCMC methods, and is based on the BUGS model specification language. However, JAGS is designed to be interacted with entirely in R, making the bayesian inference process more seamless, using the R package `rjags`. The `runjags` is an optional package that allows for the JAGS model script to be inputted as a string, and run using the `run.jags()` function. JAGS is also built on C++, which makes running the MCMC sampling process very fast. Sampling diagnostics can be visualized through the `coda` package. There is plenty of documentation and support available for JAGS, which is still maintained (JAGS manual).

@@

Nimble allows for bayesian computation within R, like JAGS and OpenBUGS, but is also made for more flexible statistical modeling. It uses C++ under the hood to compile models and algorithms. Using the `nimble` package, Nimble can perform MCMC bayesian sampling using common algorithms, but also allows for other types of algorithms to be loaded or custom-built (Michaud). It can also compile other types of algorithms without the BUGS syntax, written in a familiar fashion for R users.

For the purpose of bayesian sampling using MCMC, Nimble is used very similarly to JAGS and OpenBUGS in practicality, as its model specification language is close to the same. The flexibility of Nimble is not needed in most cases when doing standard bayesian MCMC methods. Nimble can load and use BUGS model and data files if desired. The default sampling method in Nimble is Gibbs sampling, but other algorithms are built-in as well, like Slice sampling, reversible jump MCMC, and Sequential Monte Carlo (Nimble Manual). Because the model has to be compiled first into C++ before sampling each time, this can increase the total computation time as compared to other sampling methods. There is a very accessible user manual for Nimble available online.

## Stan

Stan is a probabilistic programming language in-and-of itself, used for bayesian statistical modeling. Stan has two key differences from the BUGS-based implementations: it has its own language and syntax, and it uses a different MCMC algorithms. Stan has interfaces for R, Python, and the command line.

The Stan language is designed to be more flexible and expressive than the BUGS-type languages. Explicit variable declarations are required in Stan, and, unlike the BUGS models, and statements are executed in the order that they appear in the model file. It also supports more programming techniques, such as conditional statements using if-else syntax (Carpenter).

Stan uses Hamiltonian Monte Carlo sampling methods, which ideally is more efficient and robust than Gibbs or Metropolis-Hastings strategies. In some cases, this can reduce the computational time for many types of statistical models. Like Nimble, Stan model code is translated and compiled to C++ for faster use. In addition, Stan conveniently gives the user the option to store a compiled model in a file, which allows for repeated use without the need to re-compile each time. Stan supports both the Euclidean HMC sampler, and the no-U-turn sampler (NUTS), which is the default method (Carpenter). NUTS automatically adapts the number of leapfrog steps in the warmup phase and does not require user-specified tuning parameters (Stan User Manual).

In R, Stan is used through the `rstan` package, which allows for data to be loaded through normal R data structures. The biggest practical difference in using Stan compared to other packages is in the model specification process. Stan requires a model object in the form of a .stan file. This file, either created through a character string in R or in a separate text file, can be used by the `stan()` function, which takes the file along with the data to perform the sampling. There is a very robust set of manuals for Stan users, including many papers written on the subject.

## greta

Greta, a more recent addition to the list of bayesian computational implementations, is made specifically for easy use in R. The model and data specification process is intended to be as straightforward as possible for R users, who need not be familiar with other bayesian inference tools. Not only does it utilize Hamiltonian Monte Carlo methods for sampling, but it also uses Google TensorFlow,a machine learning platform from Google, to perform its calculations. This allows for more flexible use of CPUs and GPUs. Greta offers a model visualization tool that does not exist in any of the other Bayesian samplers, in that it can provide a graphical representation of the specified model, similar to a directed acyclical graph (greta Manual).

As of the writing of this, greta is still in need of lots of development, and lacks an appropriate amount of documentation and support for users, compared to the other implementations mentioned previously. The sampling process can slow compared to the other bayesian methods when using identical sampling configurations, especially for simple models. There are also several types of models for which greta seems unable to properly specify at the moment.

@@

### OpenBUGS

@@

model {
  y ~ dbin(p,n)
  p ~ dbeta(1,1)
}
```

@@

### JAGS

@@

### Nimble

@@

### Stan

Running models in Stan requires a .stan file which contains the model script code, written in the unique Stan syntax. This can either be a seperate file, or can be inputted as a character string. It is generally recommended to use a seperate .stan file, espcially when using RStudio, which can do syntax highlighting and point out errors in Stan code.

Here is the .stan file for this example, which is saved as "stan_file.stan". Each Stan file contains a data block, parameters block, and model block at a minimum. The data block initializes each data variable, the parameters block initializes any parameters to monitor, and the model block contains more familiar looking model script code.

@@

model {
  y ~ binomial(n,p);
  p ~ beta(1,1);
}
```

@@

### greta

Model specification in greta is more intuitive for R users, but requires the extra step of converting existing data objects into greta arrays, using the `as_data()` function from the `greta` package. 
```{r}
## convert data into greta arrays
y <- greta::as_data(y)
n <- greta::as_data(n)
```

@@

When using Markov Chain Monte Carlo methods, the number of independent Markov chains needs to be specified. It is recommended to use several chains to explore the space of the target distribution, as a single chain may not fully traverse the entire region on its own. For some simpler models that don't have irregular features in the target distribution, using one chain can suffice. Each sampler has an argument in its function call that specifies the number of chains. Some of the samplers have a default value if the number of chains isn't given.

@@

The number of total draws that each Markov chain will collect is a very important setting to get right. If each chain is not set up to run long enough, the parameter space may not be fully explored and the estimates will be inaccurate. Checking convergence statistics can be helpful in determining if the number of iterations is sufficient. An argument specifiying the number of iterations for each chain can be inputted in the function call for each sampler.

@@

Since our inferences are based on the assmumption that our draws are from the target distribution, it is wise to discard the samples at the beginning of each chain, since it takes time for the Markov chains to converge to the target distribution. The terms "warmup" and "burn-in" are often used to indicate the number of iterations that we will discard at the beginning of each chain when calculating statistics about our posterior distribution. While "warmup" is sometimes used interchangeably with "burnin", the term is usually associated with Hamiltonian Monte Carlo samplers, indicating the number of iterations that the sampler will spend adapting and fine-tuning parameters before collecing official draws. Gibbs Samplers usually use "burnin" to indicate the number of initial draws that will be discarded at the beginning of each chain before keeping track of the draws.

We can choose these values for warmup/burnin in the function call for each sampler. Some samplers (greta, JAGS, Nimble) treat this argument as a number of iterations to run before starting to track the sampling results, but others (Stan, OpenBUGS) have this be the actual number of iterations to discard out of the total number of specified iterations, so it is important to know which designation a sampler uses. For example, if one were to have 5000 iterations with 1000 burnin iterations, in JAGS this would lead to 1000 discared iterations, followed by 5000 draws, but in OpenBUGS, there would 5000 total draws, with the first 1000 discared, leaving 4000 posterior draws for each chain.

@@

When describing a bayesian model through code, there are instances where slightly different specifications are needed based on the sampler being used. Both in model and data specification, the user needs to understand the intricacies of the specific sampling tool being used in order to get the analysis to run properly. This section provides some helpful tips for implementing models in each of the different samplers being examined.
